Mergeable Types for Highly-Available Replicated Data

### 2017-02-25 paper/mergeable-types.tex

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, merge function has to enforce the relationship
explicitly. In other words, the result of merging of a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
must also exist morphisms $P'$ and $Q'$, the map both $B$ and $C$ to
$D:t$. This intuition is expressed visusally in
Fig.~\ref{fig:pushouts}.

The three-way merge function \C{merge} can be defined
straightforwardly in terms of the function $M$ described above. Thus,
for a mergeable type $t$, we have the guarantee that a sequence of
values read by a thread from its branch ``makes sense'' as per the
data type semantics:

\begin{theorem} [\bfseries Branch-local consistency]
Let $t$ be a mergeable type, and let $H$ be a legal branching history
over values of type $t$. For every pair of values $v_1:t$ and $v_2:t$
along a branch $b$ in $H$, if $\under{H}{v_1 \preceq v_2}$, then there
exists a morphism $P:t\rightarrow t$ such that $v_2 = P(v_1)$.
\end{theorem}

>>>>

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, merge function has to enforce the relationship
explicitly. In other words, the result of merging of a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type}
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
must also exist morphisms $P'$ and $Q'$, the map both $B$ and $C$ to
$D:t$. This intuition is expressed visusally in
Fig.~\ref{fig:pushouts}.

For a mergeable type $t$, we have the guarantee that a sequence of
values read by a thread from its branch ``makes sense'' as per the
data type semantics:

\begin{theorem} [\bfseries Branch-local consistency]
\label{thm:branch-consistency}
Let $t$ be a mergeable type, and let $H$ be a legal branching history
over values of type $t$. For every pair of values $v_1:t$ and $v_2:t$
along a branch $b$ in $H$, if $\under{H}{v_1 \preceq v_2}$, then there
exists a morphism $P:t\rightarrow t$ such that $v_2 = P(v_1)$.
\end{theorem}

Def.~\ref{def:mergeable-type} specifies the merge function ($M$)
slightly differently; instead of returning a merged value like the
three-way merge function (\C{merge}), $M$ returns a pair of morphisms.
This specification is useful to preempt certain non-sensical merge
functions, and thus enforce the branch-local consistency property
(Theorem~\ref{thm:branch-consistency}). For instance, consider the
counter data type from Sec.~\ref{sec:intro}. A three-way merge
function can be defined to return a constant regardless of its
arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml} Such a merge function allows the client (a thread) to
witness the violation of the branch-local consistency property. For
example, the client may \C{sync} the counter, whose current local
value is $10$, and obtain $0$ as a result, thus witnessing a violation
of counter's monotonicity property. Fortunately,
Def.~\ref{def:mergeable-type} disallows this semantics. It is
impossible to define the function $M$ for the counter data type that
violates the monotonicity invariant, because no counter morphism
violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms. On the hand, the commutativity of the diagram in
Fig.~\ref{fig:pushouts} may not be enforced statically. It can however
be verified at run-time by checking that the morphisms returned by $M$
map the respective concurrent versions to the same value. The
operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
effect.

% While the merge function $M$ as specified by
% Def.~\ref{def:mergeable-type} has more verification value, we continue
% to use the three-way merge function \C{merge} in our examples for
% flexibility.

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it doesn't require
explicit efforts to identify $P$ and $Q$ that led to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
sophisticated data types, this process is more involved, and requires
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type-2}
A type $t$ is said to be mergeable if and only if the following
conditions hold:
\begin{itemize}
  \item There exists a function $W: t \times t \rightarrow (t
  \rightarrow t)\; \C{list}$ that accepts an object $A:t$ and its
  semantic successor $B:t$, and returns a minimal sequence of
  primitive morphisms $P^*$, whose composition maps $A$ to $B$. That
  is, $W(A,B) = P^*$, where $\denot{P^*}(A) = B$, and there does not
  exist an $R^*$ such that $\denot{R^*}(A) = B$ and $\length{R^*} <
  \length{P^*}$

  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
  \rightarrow t)$ that accepts a pair of minimal sequences of
  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
  $Q'$ and $P'$, such that for any object $A:t$, $(Q' \circ
  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
\end{itemize}
\end{definition}

\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:mergeable-string-eg} shows the edit script ...

Next, we apply operational transformation to compute $P^*$ given $Q^*$
($P^* <+ Q^*$), \emph{or} $Q^*$ given $P^*$ ($Q^* <+ P^*$); there is
no need to compute both unless verification is one of the aims.

\paragraph{}
